\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}

### 2025-05-14 paper/main.tex
fix

\begin{abstract}
Large language models like ModernBERT deliver impressive performance but face a critical deployment challenge: their substantial computational and memory demands restrict their use in resource-constrained environments. While various pruning techniques exist to address this bottleneck, most require expensive post-pruning fine-tuning—creating a paradoxical situation where model optimization itself demands significant resources. This paper introduces a novel vocabulary pruning approach for ModernBERT that eliminates the need for post-pruning recovery fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%). The proposed technique specifically targets the embedding layer—which constitutes 25.95\% of the model's parameters but remains surprisingly underexplored in compression research—through systematic token importance analysis based on TF-IDF with specialized out-of-vocabulary (OOV) token handling. Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 99.2\% of the original performance, outperforming encoder-based pruning methods at these ratios. 
Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time. This work establishes an efficient pathway for creating lightweight, task-specific transformer models through offline pre-fine-tuning optimizations rather than computationally expensive retraining.
\end{abstract}

@@

\subsection{Results}
Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
\\ \\
First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact. The TF-IDF + OOV approach performs exceptionally well, maintaining 99.2\% of the original model's average performance while reducing parameters by over 20\%.
\\ \\
Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts. For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
\\ \\
Third, pruning technique effectiveness varies by task type. Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
\\ \\
Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%). Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead. A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance. This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments. Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit. Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates. For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
\\ \\
The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).

@@

% This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact. The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 99.2\% of original model performance while reducing total parameters by 20-23\%.

>>>>

\begin{abstract}
Large language models like ModernBERT deliver impressive performance but face a critical deployment challenge: their substantial computational and memory demands restrict their use in resource-constrained environments. While various pruning techniques exist to address this bottleneck, most require expensive post-pruning fine-tuning—creating a paradoxical situation where model optimization itself demands significant resources. This paper introduces a novel vocabulary pruning approach for ModernBERT that eliminates the need for post-pruning recovery fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%). The proposed technique specifically targets the embedding layer—which constitutes 25.95\% of the model's parameters but remains surprisingly underexplored in compression research—through systematic token importance analysis based on TF-IDF with specialized out-of-vocabulary (OOV) token handling. Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 97.6\% of the original performance, outperforming encoder-based pruning methods at these ratios. 
Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time. This work establishes an efficient pathway for creating lightweight, task-specific transformer models through offline pre-fine-tuning optimizations rather than computationally expensive retraining.
\end{abstract}

@@

\subsection{Results}
Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
\\ \\
First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact. The TF-IDF + OOV approach performs exceptionally well, maintaining 97.6\% of the original model's average performance while reducing parameters by over 20\%.
\\ \\
Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts. For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
\\ \\
Third, pruning technique effectiveness varies by task type. Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
\\ \\
Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%). Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead. A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance. This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments. Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit. Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates. For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
\\ \\
The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).

@@

% This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact. The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 97.6\% of original model performance while reducing total parameters by 20-23\%.
